---
title: "Example Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(susR)      
library(dplyr)
library(ggplot2)
library(stringr)
library(gganimate)
library(plotly)
```

This script demonstrates how to use the susR package to retrieve data on population size and mean age in Slovak districts from the Slovak Statistical Office API, specifically from tables om7011rr and om7005rr.

It includes:
 1) Retrieving list of district codes.
 2) Preparing parameters for API call.
 3) Retrieving the data with get_susr_data().
 4) Merging data.
 5) Performing a basic analysis.
 6) Plotting the results.


## Construct parameters for get_susr_data

The table_codes are "om7013rr" and "om7005rr".

The dimension parameters must be supplied in the same order as they
appear in the example href for this table. 

For om7013rr we have 3 dimension parameters in this order:

-   om7013rr_vuc: regional codes, we will supply a vector
-   om7013rr_obd: "all" years
-   om7013rr_ukaz: "IN010052" for permanently living mid-year population
-   om7013rr_poh: c("1", "2") for sex


For om7005rr we have 4 dimension parameters in this order:

-   om7005rr_vuc: regional codes, we will supply a vector
-   om7005rr_obd: "all" years
-   om7005rr_ukaz: "IN010089" for the mean age
-   om7005rr_poh: c("1", "2") for sex

Since we are interested in 79 districts, we firstly retrieve the dimension values and use the vector of values to feed `fetch_susr_data()` function. The codes are the same for both tables, so we will do it only once.

```{r}
districts <- susr_dimension_values("om7013rr",
                                   "om7013rr_vuc") |> 
  filter(str_detect(element_label, "District of"))


params <- list(
  "om7013rr", # 1st table
  list( # list of params for 1st table
    districts$element_value, # vector of district codes 
    "all",                   # all years
    "IN010052",               # population
    c("1", "2")              # sex: 1 = M, 2 = F
    
    ),
  "om7005rr", # 2nd table
  list( # list of params for 2nd table
    districts$element_value, # vector of district codes 
    "all",                   # all years
    "IN010089",              # mean age
    c("1", "2")              # sex: 1 = M, 2 = F
    )
)
```

## Fetch data

```{r}
res <- fetch_susr_data(params)

# res is a named list with two entries
pop_df <- res[["om7013rr"]]
age_df <- res[["om7005rr"]]
```

Let's inspect the data:

```{r}
summary(pop_df)
```

```{r}
summary(age_df)
```

We already see, that there are quite some differences in regions both in population and in mean age.

## Basic data wrangling

We will adjust the raw data to be more user-friendly and follow more common conventions. Afterwards we will merge both tables into one.

```{r}
cleaned_pop_df <- pop_df |> 
  rename( # renaming to more user-friendly names
    district = om7013rr_vuc,
    year  = om7013rr_obd,
    population  = value,
    sex = om7013rr_poh
  ) |> 
  mutate(sex = ifelse(sex == "Men", 0, 1)) |> # refactoring sex
  select(district, year, sex, population)


cleaned_age_df <- age_df |> 
  rename( # renaming to more user-friendly names
    district = om7005rr_vuc,
    year  = om7005rr_obd,
    mean_age  = value,
    sex = om7005rr_poh
  ) |> 
  mutate(sex = ifelse(sex == "Men", 0, 1)) |> # refactoring sex
  select(district, year, sex, mean_age)

merged_df <- cleaned_pop_df |> 
  left_join(cleaned_age_df, by = join_by(district, year, sex), keep = FALSE)
```


## Analysis

```{r}
ggplot(merged_df |> filter(sex == 1), aes(x = year, y = mean_age, group = district, color = district)) +
  geom_line() +
  labs(title = "Mean Age Trend Over Years - Females", x = "Year", y = "Mean Age") +
  theme_minimal() +
  guides(color = "none")
```

```{r}
ggplot(merged_df |> filter(sex == 0), aes(x = year, y = mean_age, group = district, color = district)) +
  geom_line() +
  labs(title = "Mean Age Trend Over Years - Males", x = "Year", y = "Mean Age") +
  theme_minimal() +
  guides(color = "none")
```
As we can see, most of district age getting older. However there are couple of flat-ish ones disrupting this pattern. 

```{r, warning=F, fig.height=6, fig.width=12}
# Calculate population per district per year
data_bubble <- merged_df |> 
  reframe(
    .by = c(district, year),
    tot_population = sum(population),
    mean_age = round(sum(mean_age * population) / tot_population, 1)
  ) |>  
  group_by(year) |> 
  mutate(
    age_decile = ntile(mean_age, 10)  # Divide districts into deciles by mean age
  ) |> 
  ungroup()
  

# Animated interactive bubble chart
bubble_chart <- plot_ly(
  data_bubble,
  x = ~tot_population,
  y = ~mean_age,
  size = ~tot_population,
  sizes = c(10, 250),
  color = ~age_decile,
  frame = ~year,
  text = ~district,
  type = 'scatter',
  mode = 'markers'
) %>%
  layout(
    title = "Change over Time",
    xaxis = list(title = "Total Population"),
    yaxis = list(title = "Mean Age"),
    showlegend = FALSE
  ) |> 
  colorbar(title = "Age decile")

bubble_chart
```

