---
title: "Getting started with susR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

susR provides an R interface to the Slovak Statistical Office (SUSR) open data API, making it easier to:

*   List available tables (and their dimension codes).
*   Retrieve metadata for those tables (e.g., domain, subdomain).
*   Fetch the data itself in JSON-stat format.
*   Convert the data into user-friendly tibbles ready for analysis.

This vignette guides you through a basic workflow: from installing susR to fetching data.

Installation
If you haven't already installed susR, you can do so from GitHub:

```{r eval=FALSE}
devtools::install_github("Arnold-Kakas/susR")
```

Then load it:

```{r}
library(susR)
```

## Listing Available Tables

There are two options how to retrieve available tables.

First is calling `susr_domains()` functions which reads saved list of available tables, their names, domain and subdomains:

```{r}
susr_domains_table <- susr_domains()
head(susr_domains_table)
```

Second is to call the SUSR API:

```{r}
susr_tables <- susr_tables()
head(susr_tables)
```

This returns a tibble with columns like table_code, href, dimension_names, label, and update.

If you only want certain domains or subdomains, you can do it by passing one (or more) domains and/or subdomains from `susr_domains()` result:

```{r}
# Example: filter by domain = "Demographic and social statistics"
domain_tables <- susr_tables(domain = "Demographic and social statistics")

dim(domain_tables)
```

```{r}
subdomain_tables <- susr_tables(domain = "Housing")

dim(subdomain_tables)
```

```{r}
multiple_subdomain_tables <- susr_tables(domain = c("Housing", "Energy"))

dim(multiple_subdomain_tables)
```

You can pass parameter `long = TRUE` in order to fetch data in long format where each row contain only one dimension.

```{r}
subdomain_tables_long <- susr_tables(domain = "Housing",
                                     long = TRUE)

dim(subdomain_tables_long)
```

## Discovering Dimensions

You can see what valid dimension values by using `susr_dimension_values()`.
```{r}
dims <- susr_dimension_values(table_code = "bv3001rr",
                              dimension_code = "bv3001rr_voda")

dims
```

You can map this function to `susr_tables()` output in long format.

```{r}
library(dplyr)
library(purrr)
library(tidyr)

# 1) Get all tables in the "Energy" domain, in long format
energy_long <- susr_tables(domain = "Energy", long = TRUE)

# Inspect the columns:
# - table_code
# - dimension_code
# - plus other metadata (domain, subdomain, label, href, etc.)

# 2) Map susr_dimension_values() across each row:
#    We'll create a new list-column 'dim_info' that stores the dimension values
energy_long_dim <- energy_long |> 
  mutate(
    dim_info = map2(table_code, dimension_code, ~ susr_dimension_values(.x, .y))
  )

# Now `energy_long_dim$dim_info` is a list of tibbles/data frames (one for each dimension).
# Each entry holds columns like:
#   dimension_code, dimension_label, dimension_note, element_code, element_label, etc.

# 3) (Optional) Unnest into a single combined data frame
#    If you'd like a tall data frame where each row is a unique dimension element:
full_dim_info <- energy_long_dim |> 
  unnest(cols = dim_info, names_sep = "dim_")

# `full_dim_info` now contains columns:
#   class, href, table_code, label, update, dimension_code, 
#   (optionally) domain/subdomain, 
#   and the un-nested dimension columns from susr_dimension_values().

head(full_dim_info)
```


## Fetching Data

Once you know the table code and dimension segments, use `fetch_susr_data()`. If you want to fetch data from multiple tables multiple pairs of table codes and dimension lists:

```{r}
params <- list(
  "np3106rr",
  list("SK021", c("2016","2017","2018"), "E_PRIEM_HR_MZDA", "all"),
  "as1001rs",
  list("all", "all", "all")
)

multi_res <- fetch_susr_data(params)

names(multi_res)

class(multi_res$np3106rr)
```

Each element in multi_res is a tibble for given table code.

## Example Analysis

Once you have the data, use dplyr and ggplot2 (or anything else) to analyze:

```{r out.width="100%", fig.width=8, fig.height=6}
library(ggplot2)

p <- multi_res$as1001rs |>
  filter(as1001rs_ukaz == "Pre-productive population (aged 0 -14 years)",
         as1001rs_poh == "Total") |>
  ggplot() +
  geom_line(aes(x = as1001rs_rok, y = value, group = 1)) +
  labs(title = "Pre-productive population (aged 0 -14 years)",
       x = "Year",
       y = NULL) +
  theme_minimal()


p
```
